---
title: "SYS 5581 Project - Extract, Transform, and Load Data"
author: "Nick Coronato"
date: "Version of `r Sys.Date()` | Due 2021-02-22"
output: 
    pdf_document:
      keep_md: true
---

```{r load libraries, include = FALSE}
# Call the appropriate libraries.
library(tidyverse)
library(lubridate)
library(tsibble)
library(data.table)
```

**Step 1: Identify a Time Series data set that you want to work with**

For this project, I will be analyzing a set of exercise data for 186 patients.

**Step 2: Acquire the data from its source location, reproducibly.**

For this project, my data is stored on my local machine

*Note: Ideally the data will be stored at and read from a Github repository. (Note: permission was granted to use this data, and no identifiable patient information is included in the raw data.)*

```{r Read in data, message = FALSE}
url = 'https://github.com/uva-eng-time-series-sp21/coronato-nicholas/blob/main/CPET_raw_data.csv'

# To download data from Github, need to access the "Raw" version:
url <- "https://raw.githubusercontent.com/uva-eng-time-series-sp21/coronato-nicholas/main/CPET_raw_data2.csv?token=AAWINTXAEGBKO2L3X6AQOSDAF63HK"

# Alas, the token apparently expires after one download. So we'll load the file from local copy.

(CPET_raw  <-  read_delim("CPET_raw_data2.csv",",",
                         col_types = cols(.default = col_character(),
                                          "HR" = col_double(),
                                          "VO2" = col_double(),
                                          "VO2/kg" = col_double(),
                                          "VCO2" = col_double(),
                                          "RQ" = col_double(),
                                          "VE" = col_double(),
                                          "VE/VO2" = col_double(),
                                          "VE/VCO2" = col_double(),
                                          "Work" = col_double(),
                                          "PetO2" = col_double(),
                                          "PetCO2" = col_double(),
                                          "VEO22"= col_double(),
                                          "TMSPD" = col_double(),
                                          "TMELV" = col_double()
                         )))
```

**Step 3: Organize your data into a *tidy* data frame.**

Organize by taking out the non-useful variables.

```{r Organize data into a tidy data frame}
CPET_raw <- select(CPET_raw, -LocalTime, -FE02, -FECO2,	-RER,	-RR,	-METS,	-Vtex,	-Vtin, -Source, -TypeUser, -Summary)

CPET_raw %>% filter(PatientId != SessionId)

CPET_raw %>% 
  select_if(function(x) !all(is.na(x)))

```

Make a new variable called Index so that each observation is individually identifiable, i.e. Session 1, Obs 1

```{r Generate index variable}

#load package
require(data.table)

#  Turn data.frame into a data.table
CPET_ts2 <- data.table( CPET_raw )

#  Get running count by SessionId
CPET_ts2[ , Index := 1:.N , by = c("SessionId") ]

```

Make Index variable to be a two digit readout (i.e. 01, 02, ...)

```{r Index variable to two digits, include = FALSE}

#Make Index variable to be a two digit readout (i.e. 01, 02, ...)
CPET_ts2$Index <- sprintf("%02d", as.numeric(CPET_ts2$Index))

```

Convert time column into a more usable value (seconds instead of HH:MM:SS)

```{r Convert time variable to HHMMSS, echo = FALSE}
#Convert time column into a more usable value (seconds instead of HH:MM:SS)

CPET_ts <- CPET_ts2 %>% 
  mutate(NewTime = hms(Time)) %>% 
   unite(identifier, c(SessionId, Index), sep = ".", remove = FALSE) %>% 
  select(PatientId, SessionId, identifier, NewTime, everything())

```

This can be used to create a dataframe of HR over time, per patient session.

```{r Create a HR_by_patient object}

#This can be used to create a df of HR over time, per patient session

(CPET_ts %>%
  group_by(SessionId, NewTime) %>%
  summarise(HR) -> HR_by_patient)
```

This chunk is for example purposes; ggplot of Patient 1's heart rate over time.

```{r Example ggplot HR over time, echo = FALSE}

# This chunk is for example purposes; ggplot of Patient 1's heart rate over time.

CPET_ts %>%
  filter(SessionId == 1) %>%
  ggplot(mapping = aes(x = Time, y = HR)) +
  geom_point() +
  ggtitle("Patient 1 Heart Rate Over Time")

```

Generate and print the tsibble.

```{r Generate tsibble, echo = FALSE}

#Generate tsibble

CPET_tsbl <- as_tsibble(CPET_ts, key = identifier, index = NewTime)

#Print tsibble

print(CPET_tsbl)


```
```{r}
CPET_tsbl %>%
  filter(PatientId %in% 1:3) %>%
  ggplot(mapping = aes(x = Time, y = HR)) +
  geom_line() +
  ggtitle("Patient 1 Heart Rate Over Time")

```
```{r run diagnostics on tsibble}
interval(CPET_tsbl)

is_regular(CPET_tsbl)

is_ordered(CPET_tsbl)

has_gaps(CPET_tsbl) %>% filter(.gaps == TRUE)
```
# Redo code

```{r}

# Confirm that values of PatiendId and SessionId are always identical:
CPET_raw %>% filter(PatientId != SessionId)  # Returns zero rows.

CPET_raw %>%
  select(-SessionId) %>%                        # Drop redundant SessionId field
  select_if(function(x) !all(is.na(x))) %>%     # Drop fields containing only NA values
  mutate(TestLevel = as.factor(TestLevel)) %>%  # Convert data type from text to factors
  mutate(Time = lubridate::as.duration(hms(Time))) %>%  # Convert Time to duration type 
  mutate(PatientId = as.integer(PatientId)) %>% # Convert PatientId to integer type
  select(PatientId, Time, everything()) -> CPET_rare

# Reference: https://r4ds.had.co.nz/dates-and-times.html#durations 

CPET_rare %>% arrange(desc(PatientId), desc(Time))

CPET_rare %>% 
  group_by(Time) %>% 
  summarize(N = n())
```
```{r try to convert tibble to a tsibble}
# CPET_rare %>% as_tsibble(index = Time, key = PatientId)
# The above line throws an error: Says there are duplicates.
# Check for duplicates:
(CPET_rare %>% 
  duplicates(index = Time, key = PatientId) -> CPET_dupes)

# Examine these duplicates in the original CPET_raw table
CPET_raw %>%
  filter(PatientId %in% CPET_dupes$PatientId & lubridate::as.duration(hms(Time)) %in% CPET_dupes$Time)

# Need somehow to resolve or drop duplicates. 
# This code will drop the second record in each duplicated group:
CPET_dupes %>%
      group_by(PatientId, Time) %>%
      slice(1)

CPET_rare %>%
      group_by(PatientId, Time) %>%
      slice(1) %>%
      ungroup() -> CPET_medium

# Reference: https://stackoverflow.com/questions/22959635/remove-duplicated-rows-using-dplyr
```

```{r convert tibble to a tsibble}
(CPET_medium %>% as_tsibble(index = Time, key = PatientId) -> CPET_tbl_ts)
```

```{r run diagnostics on new tsibble}
interval(CPET_tbl_ts)

is_regular(CPET_tbl_ts)

is_ordered(CPET_tbl_ts)

has_gaps(CPET_tbl_ts) 
```

```{r}
CPET_tbl_ts %>%
  filter(PatientId == 1) %>%
  ggplot(mapping = aes(x = Time, y = HR)) +
  geom_line() + geom_point() +
  ggtitle("Patient 1 Heart Rate Over Time") + xlab("Time (seconds)") + ylab("Heart rate (s^-1)")
```
```{r}
CPET_tbl_ts %>%
  filter(PatientId %in% 1:4) %>%
  feasts::autoplot(HR) + xlab("Time (seconds)") + ylab("Heart rate (s^-1)")
```

```{r Bin times into twenty-second intervals}
# Reference: https://community.rstudio.com/t/how-to-bin-a-time-column-in-r/44867/2
CPET_medium %>%
  mutate(Time_20s = hms::round_hms(Time, secs = 20)) %>%
  group_by(PatientId, Time_20s) %>%
  summarise(HR20 = mean(HR)) %>%
  ungroup() %>%
  as_tsibble(index = Time_20s, key = PatientId)
```

```{r}
CPET_medium %>% filter(HR == 0)
```
It looks like some of the patients have a recorded heart rate of zero. Either there is another round of data cleaning to do, or [some other procedue should be followed](https://www.youtube.com/watch?v=QcbR1J_4ICg).

