---
title: "SYS 5581 Project - Extract, Transform, and Load Data - 2nd Attempt"
author: "Nick Coronato"
date: "Version of `r Sys.Date()` | Due 2021-03-01"
output: pdf_document
---


```{r load libraries, include = FALSE}
# Call the appropriate libraries.
library(tidyverse)
library(lubridate)
library(tsibble)
library(data.table)
library(fable)


```

**Step 1: Identify a Time Series data set that you want to work with**

For this project, I will be analyzing a set of airline passenger data from the Bureau of Transportation Statistics (BTS). The data set includes air travel data from September 2018 - August 2020 (24 months).



**Step 2: Acquire the data from its source location, reproducibly.**

For this project, my data is stored on my local machine. The file name is "2year_t100_data.csv"

Data was obtained from the BTS website: https://www.transtats.bts.gov/DL_SelectFields.asp?gnoyr_VQ=FIL&QO_fu146_anzr=Nv4%20Pn44vr45

*Note: Ideally the data will be stored at and read from a Github repository. *

```{r Read in data, message = FALSE}

t100_raw  <-  read_delim("2year_t100_data.csv",",",
                         col_types = cols(.default = col_character(),
                                          "YEAR" = col_integer(),
                                          "MONTH" = col_integer(),
                                          "UNIQUE_CARRIER" = col_factor(),
                                          "PASSENGERS" = col_integer(),
                                          "DISTANCE" = col_integer(),
                                          "ORIGIN_AIRPORT_ID" = col_factor(),
                                          "ORIGIN_STATE_NM" = col_factor(),
                                          "DEST_AIRPORT_ID" = col_factor(),
                                          "DEST_STATE_NM" = col_factor(),
                                          "SEATS" = col_integer(),
                                          "AIR_TIME" = col_integer()
                                                  ))
                                          
```

**Step 3: Organize your data into a *tidy* data frame.**

Organize by taking out the non-useful variables. Make another new variable called RPM (revenue passenger miles), which is simply Passengers X Miles for each observation. Put the Index variable in the first column (eventually it will be 4dig year, 2dig month). Put the Key variables next (unique_carrier, Passengers, Distance, RPM). Then all the other (possibly) relevant variables.

```{r Organize data into a tidy data frame}
t100_raw %>% 
  mutate(., RPM = PASSENGERS * DISTANCE) %>% 
  select(YEAR, MONTH, UNIQUE_CARRIER, PASSENGERS, DISTANCE, RPM, -ORIGIN_AIRPORT_ID, -ORIGIN_CITY_NAME, -ORIGIN_STATE_NM, -DEST_AIRPORT_ID, -DEST_CITY_NAME, -DEST_STATE_NM, SEATS, AIR_TIME, -PAYLOAD, -ORIGIN_COUNTRY, -ORIGIN_COUNTRY_NAME, -DATA_SOURCE) -> t100_raw

```


Make a new variable called YRMO that concatenates Year and Month into a YR-MO format. Make it a time series column.

```{r Generate index variable for tsibble, message = FALSE}

# Put together YEAR and MONTH into a single column called YRMO
t100_raw %>% 
  mutate(YRMO = paste(t100_raw$YEAR, t100_raw$MONTH)) ->t100_raw

# Convert YRMO to a time series object column (year month) and re-arrange
t100_raw %>% 
  mutate(YRMO = yearmonth(YRMO))  %>% 
  select(YRMO, everything(),-MONTH, -YEAR)-> t100_raw2

# Create summary values for Passengers, RPM, Distance, Seats, and Air_time that will prevent us from having duplicates. These stats are now rolled up monthly by airline 
t100_raw2 %>% 
  group_by(YRMO, UNIQUE_CARRIER) %>%
  summarize(TotalPax=sum(PASSENGERS), TotalRPM = sum(RPM), TotalDistance = sum(DISTANCE), TotalSeats = sum(SEATS), TotalAIR_TIME = sum(AIR_TIME))  -> t100_raw3

# convert to tsibble
as_tsibble(t100_raw3, index = YRMO, key = UNIQUE_CARRIER) -> t100_ts_tbl
  
```



Some quality control checks.

```{r Quality Control of tsibble}

#Check that each column got the right Type.

head(t100_ts_tbl)

```


This chunk is for example purposes; an unrefined ggplot of passnegers over time.

```{r Example ggplot passengers over time, warning = FALSE, echo = FALSE}

# This chunk is for example purposes


# t100_ts_tbl %>% 
#   filter(UNIQUE_CARRIER == c("UA","AA","DL")) %>% 
#   autoplot(t100_ts_tbl,aes(YRMO,TotalPax, color = UNIQUE_CARRIER))

t100_ts_tbl %>%
  filter(UNIQUE_CARRIER == c("AA","UA","DL")) %>%
  ggplot(mapping = aes(x = YRMO, y = TotalPax, color = UNIQUE_CARRIER)) + geom_point() + ggtitle("3 Major Airlines Monthly Passengers")

```


Generate and print the tsibble.

```{r Generate tsibble, echo = FALSE}

#Print tsibble

print(t100_ts_tbl)


```

