---
title: "SYS 5581 Project - Extract, Transform, and Load Data"
author: "Nick Coronato"
date: "Version of `r Sys.Date()` | Due 2021-02-22"
output: pdf_document
---


```{r load libraries, include = FALSE}
# Call the appropriate libraries.
library(tidyverse)
library(lubridate)
library(tsibble)
library(data.table)
```

**Step 1: Identify a Time Series data set that you want to work with**

For this project, I will be analyzing a set of exercise data for 186 patients.


**Step 2: Acquire the data from its source location, reproducibly.**

For this project, my data is stored 
#on my local machine

*Note: Ideally the data will be stored at and read from a Github repository. (Note: permission was granted to use this data, and no identifiable patient information is included in the raw data.)*

```{r Read in data, message = FALSE}
url = 'https://github.com/uva-eng-time-series-sp21/coronato-nicholas/blob/main/CPET_raw_data.csv'

(CPET_raw  <-  read_delim("CPET_raw_data2.csv",",",
                         col_types = cols(.default = col_character(),
                                          "HR" = col_double(),
                                          "VO2" = col_double(),
                                          "VO2/kg" = col_double(),
                                          "VCO2" = col_double(),
                                          "RQ" = col_double(),
                                          "VE" = col_double(),
                                          "VE/VO2" = col_double(),
                                          "VE/VCO2" = col_double(),
                                          "Work" = col_double(),
                                          "PetO2" = col_double(),
                                          "PetCO2" = col_double(),
                                          "VEO22"= col_double()
                         )))
                                          
```

**Step 3: Organize your data into a *tidy* data frame.**

Organize by taking out the non-useful variables.

```{r Organize data into a tidy data frame}
CPET_raw <- select(CPET_raw, -LocalTime, -FE02, -FECO2,	-RER,	-RR,	-METS,	-Vtex,	-Vtin, -Source, -TypeUser, -Summary)

```


Make a new variable called Index so that each observation is individually identifiable, i.e. Session 1, Obs 1

```{r Generate index variable}

#load package
require(data.table)

#  Turn data.frame into a data.table
CPET_ts2 <- data.table( CPET_raw )

#  Get running count by SessionId
CPET_ts2[ , Index := 1:.N , by = c("SessionId") ]

```


Make Index variable to be a two digit readout (i.e. 01, 02, ...)

```{r Index variable to two digits, include = FALSE}

#Make Index variable to be a two digit readout (i.e. 01, 02, ...)
CPET_ts2$Index <- sprintf("%02d", as.numeric(CPET_ts2$Index))

```



Convert time column into a more usable value (seconds instead of HH:MM:SS)

```{r Convert time variable to HHMMSS, echo = FALSE}
#Convert time column into a more usable value (seconds instead of HH:MM:SS)

CPET_ts <- CPET_ts2 %>% 
  mutate(NewTime = hms(Time)) %>% 
   unite(identifier, c(SessionId, Index), sep = ".", remove = FALSE) %>% 
  select(PatientId, SessionId, identifier, NewTime, everything())

```



This can be used to create a dataframe of HR over time, per patient session.

```{r Create a HR_by_patient object}

#This can be used to create a df of HR over time, per patient session

(CPET_ts %>%
  group_by(SessionId, NewTime) %>%
  summarise(HR) -> HR_by_patient)
```


This chunk is for example purposes; ggplot of Patient 1's heart rate over time.

```{r Example ggplot HR over time, echo = FALSE}

# This chunk is for example purposes; ggplot of Patient 1's heart rate over time.

CPET_ts %>%
  filter(SessionId == 1) %>%
  ggplot(mapping = aes(x = Time, y = HR)) +
  geom_point() +
  ggtitle("Patient 1 Heart Rate Over Time")

```


Generate and print the tsibble.

```{r Generate tsibble, echo = FALSE}

#Generate tsibble

CPET_tsbl <- as_tsibble(CPET_ts, key = identifier, index = NewTime)

#Print tsibble

print(CPET_tsbl)


```

